Random number generation and applications on GPU’s in R(?)

### 2021-10-26 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
chol fixed ithink
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2930 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.100   0.004   0.459
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   4.640   0.304   4.939
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

month_gpu<-vclMatrix(month,type="integer")
result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,128))
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410644
result_month$p.value
## [1] 0.4042541
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,128)))
##    user  system elapsed 
##   0.709   0.004   2.326
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
## [1] 0.4039421
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  10.248   0.000  10.242
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for 1015808 simulations, has a p-value of 0.4035.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,128))
result_weekgpu$simNum
## [1] 10027008
result_weekgpu$counts
## [1] 1242
result_weekgpu$p.value
## [1] 0.0001239652
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1223 cases and a p-value around 0.0001224705. We summarized the results in Table \ref{tab:summary}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = c(256,128)))
##    user  system elapsed 
##   0.607   0.012  10.764
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 90.1 seconds, while \fct{fisher.sim} takes about 1.90 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=10010624)$p.value
## [1] 0.0001263657
system.time(fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
##  62.792   0.036  62.770
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001208, which should require a larger number of simulations to get a more accurate p-value.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
myParamsBatch
##      shape range variance nugget anisoRatio anisoAngleRadians
## [1,]  1.25  0.50      1.5      0          1         0.0000000
## [2,]  2.15  0.25      2.0      0          4         0.4487990
## [3,]  0.55  1.50      2.0      0          4         0.4487990
## [4,]  2.15  0.50      2.0      0          4        -0.4487990
## [5,]  2.15  0.50      2.0      0          2         0.7853982
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
\end{verbatim}
\end{kframe}
\end{knitrout}
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
## [1] 24000  4800
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(128,64), Nlocal=c(16,4))
\end{verbatim}
\end{kframe}
\end{knitrout}
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal= c(128, 8),
                     Nlocal= c(32, 8),
                     NlocalCache=1000)
\end{verbatim}
\end{kframe}
\end{knitrout}
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streamsGpu <- createStreamsGpu(n=128*64)
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
zmatGpu = rnorm(c(nrow(maternCov),2), streams=streamsGpu, Nglobal=c(128,64),
                 type = theType)
\end{verbatim}
\end{kframe}
\end{knitrout}
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)

>>>>

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $512 \times 128$ work-items is around 120 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.144   0.004   0.189
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   6.762   0.868   7.624
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

month_gpu<-vclMatrix(month,type="integer")
result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,128))
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410644
result_month$p.value
## [1] 0.4042541
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 1015808, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.40375, which is close enough to the p-value from \fct{stats::fisher.test}: 0.40380.  
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(fisher.sim(month_gpu, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = c(256,128)))
##    user  system elapsed 
##   0.720   0.001   0.721
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 0.32 seconds, \pkg{gpuRandom} accounts for 2.2\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
## [1] 0.4035335
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  14.706   0.004  14.730
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for 1015808 simulations, has a p-value of 0.4035.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,128))
result_weekgpu$simNum
## [1] 10027008
result_weekgpu$counts
## [1] 1242
result_weekgpu$p.value
## [1] 0.0001239652
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1223 cases and a p-value around 0.0001224705. We summarized the results in Table \ref{tab:summary}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = c(256,128)))
##    user  system elapsed 
##   4.585   0.001   4.582
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 90.1 seconds, while \fct{fisher.sim} takes about 1.90 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=10010624)$p.value
## [1] 0.0001292627
system.time(fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
##  90.434   0.196  90.561
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001208, which should require a larger number of simulations to get a more accurate p-value.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
myParamsBatch
##      shape range variance nugget anisoRatio anisoAngleRadians
## [1,]  1.25  0.50      1.5      0          1         0.0000000
## [2,]  2.15  0.25      2.0      0          4         0.4487990
## [3,]  0.55  1.50      2.0      0          4         0.4487990
## [4,]  2.15  0.50      2.0      0          4        -0.4487990
## [5,]  2.15  0.50      2.0      0          2         0.7853982
paramsGpu = gpuRandom::maternGpuParam(myParamsBatch, type=theType)
\end{verbatim}
\end{kframe}
Step 3, compute Mat\'ern covariance matrices using \fct{gpuRandom::maternBatch}, the returned Mat\'ern covariance matrices %$\Sigma_1, \Sigma_2, \dots, \Sigma_5$ correspond to the 5 parameter sets respectively, the matrices 
are each of size $4800 \times 4800$ and are stacked by row in the output \code{maternCov}. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
maternCov = vclMatrix(0, nrow(paramsGpu)*nrow(coordsGpu), 
                         nrow(coordsGpu),type=theType)
dim(maternCov)
## [1] 24000  4800
gpuRandom::maternBatch(maternCov, coordsGpu, paramsGpu,
                       Nglobal=c(128,64), Nlocal=c(16,4))
\end{verbatim}
\end{kframe}
Step 4, the first argument \code{maternCov} in \fct{gpuRandom::cholBatch} specifies the matrix object to take Cholesky decomposition. Unit lower triangular matrices $L_i$'s are returned and stacked by row in \code{maternCov}, \code{diagMat} stores the diagonal values of each $D_i$ in a row, for example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $L_i$ is $n \times n$, and each batch $D_i$ is $1 \times n$.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
diagMat = vclMatrix(0, nrow(paramsGpu), ncol(maternCov), type = theType)
gpuRandom::cholBatch(maternCov, diagMat, numbatchD=nrow(myParamsBatch),
                     Nglobal= c(128, 8),
                     Nlocal= c(32, 8),
                     NlocalCache=1000)
## 
## #pragma OPENCL EXTENSION cl_khr_fp64 : enable
## 
## 
## #define N 4800  //dimension of matrix
## #define colStart 0  //column to start at
## #define colEnd 4800
## //internal number of columns
## #define Npad 4864
## //internal columns for matrix holding diagonals
## #define NpadDiag 4864
## #define NstartA 0
## #define NstartD 0
## //elements in internal cache
## #define Ncache 1000
## //extra rows between stacked matrices
## #define NpadBetweenMatrices 23347200
## #define maxLocalItems 256
## 
## __kernel void cholBatch(
## 	__global double *A,
## 	__global double *diag,
##  __local double*diagLocal,
##               int Nmatrix
## ){
##  const int localIndex = get_local_id(0)*get_local_size(1) + get_local_id(1);
##  const int localIndexIsZero = (localIndex==0);
##  const int NlocalTotal = get_local_size(0)*get_local_size(1);
##  local double toAddLocal[maxLocalItems];
##  int Dcol, DcolNpad;
##  int Drow, DrowBlock, Dk, Dmatrix, DmatrixBlock;
## int minDcolNcache;
##  double DL;
##  local double diagDcol;
##  int AHere, AHereDcol, AHereDrow, diagHere;
##  barrier(CLK_LOCAL_MEM_FENCE);
## 
## for(Dmatrix = get_group_id(0);
##      Dmatrix < Nmatrix;
##      Dmatrix+= get_num_groups(0)){
## 
##  diagHere = Dmatrix*NpadDiag + NstartD;
##  AHere =  Dmatrix*NpadBetweenMatrices + NstartA;
## 
##  for(Dcol = colStart; Dcol < colEnd; Dcol++) {
##   DcolNpad = Dcol*Npad;
##   AHereDcol = AHere+DcolNpad;
##   DL = 0.0;
##   diagLocal[localIndex]=0.0;
##   toAddLocal[localIndex]=0.0;
##   minDcolNcache = min(Dcol, Ncache);
##   for(Dk = localIndex; Dk < minDcolNcache; Dk += NlocalTotal) {
##     DL = A[AHereDcol+Dk];
##     diagLocal[Dk] = diag[diagHere+Dk] * DL;
##     toAddLocal[localIndex] += diagLocal[Dk] * DL;
##   }// Dk
##   for(Dk=minDcolNcache+localIndex; Dk < Dcol; Dk += NlocalTotal) {
##     DL = A[AHereDcol+Dk];
##     toAddLocal[localIndex] += diag[diagHere+Dk] * DL * DL;
##   }// Dk
## 
## // reduction on dimension 1
##   barrier(CLK_LOCAL_MEM_FENCE);
##   if( (get_local_id(1) == 0)){
##    for(Dk = 1; Dk < get_local_size(1); Dk++) {
##     toAddLocal[localIndex] +=  toAddLocal[localIndex + Dk];
##    }//for Dk
##   }//get_local_id(1) == 0
##   barrier(CLK_LOCAL_MEM_FENCE);
## 
## // final reduction on dimension 0
##   if(localIndexIsZero ){
##    for(Dk = get_local_size(1); Dk < NlocalTotal; Dk+= get_local_size(1)) {
##     toAddLocal[localIndex] +=  toAddLocal[Dk];
##    } //Dk
##    diagDcol = A[AHereDcol+Dcol] - toAddLocal[localIndex];
##    diag[diagHere+Dcol] = diagDcol;
## #ifdef diagToOne
##    A[AHereDcol+Dcol] = 1.0;
## #endif
##   }  //localIndex==0 and Dmatrix < Nmatrix
##   barrier(CLK_LOCAL_MEM_FENCE);
## 
## // off diagonals
##  for(DrowBlock = Dcol+1; DrowBlock < N; DrowBlock += get_local_size(0)) {
##   Drow = DrowBlock + get_local_id(0);
##   AHereDrow = AHere+Drow*Npad;
##   DL = 0.0;
##   if(Drow < N){
##   for(Dk = get_local_id(1); Dk < minDcolNcache; Dk+=get_local_size(1)) {
##    DL += A[AHereDrow+Dk] * diagLocal[Dk];
##   } // Dk
## 	 for(minDcolNcache + get_local_id(1); Dk < Dcol; Dk+=get_local_size(1)) {
##     DL += A[AHereDrow+Dk] * diag[diagHere+Dk] * A[AHereDcol+Dk];
##   }// Dk
##    }//Drow < N
##   toAddLocal[localIndex] = DL;
## 
##   // local reduction
##   barrier(CLK_LOCAL_MEM_FENCE);
##   if( (get_local_id(1) == 0) & (Drow < N)){
##    DL = toAddLocal[localIndex];
##    for(Dk = 1; Dk < get_local_size(1); Dk++) {
##     DL +=  toAddLocal[localIndex + Dk];
##    }//Dk
##    A[AHereDrow+Dcol] = (A[AHereDrow+Dcol] - DL)/diagDcol;
##   }//get_local_id(1) == 0
##   barrier(CLK_GLOBAL_MEM_FENCE);
## 
##  }//DrowBlock
## barrier(CLK_GLOBAL_MEM_FENCE);
## } // Dcol loop
## 
## barrier(CLK_GLOBAL_MEM_FENCE);
## } // Dmatrix loop
## 
## barrier(CLK_LOCAL_MEM_FENCE);
## }// kernel
\end{verbatim}
\end{kframe}
% \begin{gather}
% \begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
% \end{bmatrix}
%  \rightarrow
%  \begin{bmatrix}
%   L_{1} \\ L_{2} \\L_{3} \\ \vdots
%   \end{bmatrix} \text{and}
%   \begin{bmatrix}
%   D_{1} \\ D_{2} \\D_{3} \\ \vdots
%   \end{bmatrix}
% \end{gather}
Step 5, generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{gpuRandom::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streamsGpu <- createStreamsGpu(n=128*64)
\end{verbatim}
\end{kframe}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
zmatGpu = rnorm(c(nrow(maternCov),2), streams=streamsGpu, Nglobal=c(128,64),
                 type = theType)
\end{verbatim}
\end{kframe}
Step 6, use \fct{gpuRandom::multiplyLowerDiagonalBatch} to compute $U = L * D^{(1/2)}* Z$ in batches, see the following illustration. \code{simMat} is the output matrix for $U$, \code{maternCov}, \code{diagMat}, and \code{zmatGpu} correspond to the matrices of $L$, $D$ and $Z=(Z_1, Z_2)$ respectively.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
simMat = vclMatrix(0, nrow(zmatGpu), ncol(zmatGpu), 
                   type = theType)
